Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-01 _tex/index.tex
Built site for gh-pages

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, probability, Thurstone, comparative
judgement, directed acyclic graph, structural causal models, statistical
modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Causes and effects in Dichotomous Comparative Judgments: an
information-theoretical system with plausible mechanism}
\author[1]{Jose Manuel Rivera Espejo%
\corref{cor1}%
}
 \ead{JoseManuel.RiveraEspejo@uantwerpen.be} 
\author[1]{Tine van Daal%
%
}
 \ead{tine.vandaal@uantwerpen.be} 
\author[1]{Sven De Maeyer%
%
}
 \ead{sven.demaeyer@uantwerpen.be} 
\author[2]{Steven Gillis%
%
}
 \ead{steven.gillis@uantwerpen.be}

@@

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by
\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
However, its application transcends education, as exemplified by
\citet{Boonen_et_al_2020}. The methodology has also evolved to include
multiple, as opposed to pairwise comparisons
\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
suggests that DCJ offers an alternative and efficient approach to
measurement and evaluation, characterized by its reliability and
validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
Nevertheless, despite the method's widespread use, existing literature
lacks a clear representation of the plausible mechanisms through which
DCJ data could be generated. Particularly, there is no depiction of the
complexity and the assumptions underpinning the DCJ system, nor how
different assessment factors can potentially influence the observed DCJ
outcome.

@@

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing the structural
approach \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study
develops a scientific model to clarify plausible causal assumptions and
mechanisms inherent in the DCJ system. The study then translates the
scientific model into a probabilistic statistical model. This model aims
to produce statistical estimates to draw inferences about plausible
causal relationships within the DCJ system.

Ultimately, this research provides a robust probabilistic foundation for
the statistical analysis of DCJ data, building upon Thurstone's law of
comparative judgment \citeyearpar{Thurstone_1927}. Consequently, its
findings offer valuable insights for researchers and analysts designing
and implementing DCJ experiments.

@@

This section introduces fundamental concepts in causal inference, such
as directed acyclic graphs (DAGs), structural causal models (SCMs), and
the flow of association and causation in graphs. It is not a
comprehensive description of causal inference methods. Readers
interested in deeper exploration should consult introductory papers like
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find introductory books such
as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
\citet{McElreath_2020} useful. For more advanced study, seminal
intermediate papers like \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
\citet{Hernan_et_al_2020} are recommended.

@@

Counterfactuals represent scenarios \emph{contrary to fact}, where
alternative \emph{potential} outcomes resulting from a cause are neither
observed nor observable \citep{Neal_2020, Counterfactual_2024}.
According to \citet{Pearl_et_al_2018}, counterfactuals form the
foundation of causal inference and occupy the highest level of cognitive
abstraction in the ladder of causation, followed by intervention and
association. Nevertheless, despite their abstract nature,
counterfactuals enable the development of a \emph{theory of the world}
that explains why specific causes have specific effects and what occurs
in their absence \citep{Pearl_et_al_2018}. They achieve this by
translating causal statements into counterfactual statements, that is,
statements about ``what would have happened in the world under different
circumstances.''

@@

Graph theory is a branch of mathematics focused on the study of graphs.
Graphs are mathematical structures that model pairwise relations between
objects. They can represent physical relations, such as electrical
circuits and roadways, and less tangible structures, such as ecosystems
and sociological relations. Graphs have proven useful in various fields,
including computer science, operations research, and the natural and
social sciences \citep{Gross_et_al_2018}.

@@

One key advantage of DAGs is that, regardless of complexity, they can
represent various causal structures using only five fundamental building
blocks \citep{Neal_2020, McElreath_2020}. Figure~\ref{fig-dags}
illustrates these building blocks. Figure~\ref{fig-dag_bb1} depicts two
unconnected nodes, representing an scenario where variables \(X_{1}\)
and \(X_{2}\) are not causally related. Figure~\ref{fig-dag_bb2} shows
two connected nodes, illustrating a scenario where a parent node
\(X_{1}\) exerts a causal influence on a child node \(X_{2}\).
Consequently, \(X_{2}\) is considered a \emph{descendant} of \(X_{1}\).
Figure~\ref{fig-dag_bb3} depicts a \emph{chain} (or \emph{pipe}), where
\(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In
this configuration, \(X_{1}\) is a parent node of \(X_{2}\), and
\(X_{2}\) a parent node of \(X_{3}\). Furthermore, the DAG shows that
\(X_{1}\) is an \emph{ancestor} of \(X_{3}\) and that the relationship
between these variables is entirely mediated by \(X_{2}\).
Figure~\ref{fig-dag_bb4} illustrates a \emph{fork}, where variables
\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). In this
scenario \(X_{2}\) is a parent node of both \(X_{1}\) and \(X_{3}\).
Finally, Figure~\ref{fig-dag_bb5} depicts a \emph{collider}, also known
as \emph{inmorality}, where variables \(X_{1}\) and \(X_{3}\) are
concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
\(X_{3}\) are not causally related to each other but both influence
\(X_{2}\).

@@

Given the heuristic nature of DAGs, a motivating example can illustrate
how these five fundamental building blocks can be used to construct the
causal structure of a system. This example can also help clarify
additional rules and conventions for drawing DAGs.

Consider a research problem where the causal effect of a variable \(T\)
on an outcome \(Y\) needs to be investigated. Additionally, the problem
indicates that a variable \(X\) potentially influences both \(Y\) and
\(T\). Such scenarios are not hard to imagine. For instance, \(T\) could
describe different treatments potentially influencing the binary onset
of cancer (\(Y\)) while \(X\) could denote the age of a patient. Using
an example from DCJ, {we need an example}. Figure~\ref{fig-example}
presents two graphs describing the plausible causal structure of this
system, the \emph{simplified} and the \emph{magnified} DAG.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, probability, Thurstone, comparative
judgement, directed acyclic graph, structural causal models, statistical
modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Causes and effects in Dichotomous Comparative Judgments: an
information-theoretical system of plausible mechanism}
\author[1]{Jose Manuel Rivera Espejo%
\corref{cor1}%
}
 \ead{JoseManuel.RiveraEspejo@uantwerpen.be} 
\author[1]{Tine van Daal%
%
}
 \ead{tine.vandaal@uantwerpen.be} 
\author[1]{Sven De Maeyer%
%
}
 \ead{sven.demaeyer@uantwerpen.be} 
\author[2]{Steven Gillis%
%
}
 \ead{steven.gillis@uantwerpen.be}

@@

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by
\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
However, its application transcends education, as exemplified by
\citet{Boonen_et_al_2020}. The methodology has also evolved to include
multiple, as opposed to pairwise comparisons
\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
suggests that DCJ offers an alternative and efficient approach to
measurement and evaluation, characterized by its reliability and
validity
\citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020, Bouwer_et_al_2023}.
Nevertheless, despite the method's widespread use, existing literature
lacks a clear representation of the plausible mechanisms through which
DCJ data could be generated. Particularly, there is no depiction of the
complexity and the assumptions underpinning the DCJ system, nor how
different assessment factors can potentially influence the observed DCJ
outcome.

@@

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021, Bouwer_et_al_2023},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing the structural
approach \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study
develops a scientific model to clarify plausible causal assumptions and
mechanisms inherent in the DCJ system. The study then translates the
scientific model into a probabilistic statistical model. This model aims
to produce statistical estimates to draw inferences about plausible
causal relationships within the DCJ system.

Ultimately, this study provides a robust causal and probabilistic
foundation for the statistical analysis of DCJ data, building upon
Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
Consequently, its findings offer valuable insights for researchers and
analysts designing and implementing DCJ experiments.

@@

This section introduces fundamental concepts in causal inference, such
as directed acyclic graphs (DAGs), structural causal models (SCMs), and
the flow of association and causation in graphs. The section is not a
comprehensive description of causal inference methods. Readers
interested in deeper exploration should consult introductory papers like
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find introductory books such
as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
\citet{McElreath_2020} useful. For more advanced study, seminal
intermediate papers like \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
\citet{Hernan_et_al_2020} are recommended.

@@

Counterfactuals represent scenarios \emph{contrary to fact}, where
alternative \emph{potential} outcomes resulting from a cause are neither
observed nor observable \citep{Neal_2020, Counterfactual_2024}.
According to \citet{Pearl_et_al_2018}, counterfactuals are the
foundation of causal inference and occupy the highest level of cognitive
abstraction in the ladder of causation, followed by intervention and
association. Nevertheless, despite their abstract nature,
counterfactuals enable the development of a \emph{theory of the world}
that explains why specific causes have specific effects and what occurs
in their absence \citep{Pearl_et_al_2018}. They achieve this by
translating causal statements into counterfactual statements, that is,
statements about ``what would have happened in the world under different
circumstances.''

@@

Graph theory is a branch of mathematics focused on the study of graphs.
Graphs are mathematical structures modeling pairwise relations between
objects. They can represent physical relations, such as electrical
circuits and roadways, and less tangible structures, such as ecosystems
and sociological relations. Graphs have proven useful in various fields,
including computer science, operations research, and the natural and
social sciences \citep{Gross_et_al_2018}.

@@

One key advantage of DAGs is that, regardless of complexity, they can
represent various causal structures using only five fundamental building
blocks \citep{Neal_2020, McElreath_2020}. Figure~\ref{fig-dags}
illustrates these blocks. Figure~\ref{fig-dag_bb1} depicts two
unconnected nodes, representing an scenario where variables \(X_{1}\)
and \(X_{2}\) are not causally related. Figure~\ref{fig-dag_bb2} shows
two connected nodes, illustrating a scenario where a parent node
\(X_{1}\) exerts a causal influence on a child node \(X_{2}\).
Consequently, \(X_{2}\) is considered a \emph{descendant} of \(X_{1}\).
Figure~\ref{fig-dag_bb3} depicts a \emph{chain} (or \emph{pipe}), where
\(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In
this configuration, \(X_{1}\) is a parent node of \(X_{2}\), and
\(X_{2}\) a parent node of \(X_{3}\). Furthermore, the DAG shows that
\(X_{1}\) is an \emph{ancestor} of \(X_{3}\) and that the relationship
between these variables is entirely \emph{mediated} by \(X_{2}\).
Figure~\ref{fig-dag_bb4} illustrates a \emph{fork}, where variables
\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). In this
scenario \(X_{2}\) is a parent node of both \(X_{1}\) and \(X_{3}\).
Finally, Figure~\ref{fig-dag_bb5} depicts a \emph{collider}, also known
as \emph{inmorality}, where variables \(X_{1}\) and \(X_{3}\) are
concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
\(X_{3}\) are not causally related to each other but both influence
\(X_{2}\).

@@

Given the heuristic nature of DAGs, a motivating example can help
illustrate the use of the five fundamental building blocks to construct
a system's causal structure. This example can also help to clarify
additional conventions for drawing DAGs.

Consider a research problem where the causal effect of a variable \(T\)
on an outcome \(Y\) needs to be investigated. Additionally, the problem
suggests that a variable \(X\) potentially influences both \(T\) and
\(Y\). Such scenarios are not hard to imagine. For instance, \(T\) might
represent different treatments that could affect the recovery from
cancer \(Y\), while \(X\) could denote the patient's age. Similarly, in
the context of a DCJ study like the one described by
\citet{Boonen_et_al_2020}, \(T\) could represent the duration of a
child's cochlear implant use, which might influence the child's overall
speech quality \(Y\), with \(X\) indicating the child's hearing status.
Note that these descriptions do not specify the type of data or the
functional forms of the variables, highlighting the heuristic nature of
DAGs. {(not a bad example, but I prefer one using writing skills)}

Figure~\ref{fig-example} presents two graphs illustrating the plausible
causal structure of the motivating example.
Figure~\ref{fig-dag_example1} shows the \emph{simplified} DAG. In this
graph, the variables of interest or \emph{endogenous} variables
\(V=\{T,X,Y\}\) are depicted as solid black circles, indicating that
they are observed variables. The arrows in the graph reflect the
expected direction of causal influences among the variables.
Figure~\ref{fig-dag_example2}, in contrast, depicts the \emph{magnified}
version of the DAG. This graph includes, in addition to the endogenous
variables, the \emph{exogenous} variables \(E=\{e_{T},e_{X},e_{Y}\}\).
These exogenous variables are commonly referred to as
\emph{disturbances} or \emph{errors}, because they represent factors not
explicitly modeled. They are depicted as open circles, indicating their
unobserved nature. While such exogenous variables are often omitted for
simplicity, including them in the graph can be advantageous in certain
scenarios, as they can help to highlight potential issues related to
conditioning and confounding \citep{Cinelli_et_al_2020}, concepts that
will be discussed in the next section.

\newcommand{\dsep}{\perp\!\!\!\perp}
\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}

Furthermore, a detailed examination of both figures reveals the presence
of at least four of the five fundamental building blocks.
Figure~\ref{fig-dag_example1} displays multiple descendants, evident in
pairwise relations such as \(X \rightarrow T\), \(X \rightarrow Y\), and
\(T \rightarrow Y\). It also illustrates a fork with
\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-dag_example2}
features multiple two-unconnected nodes, evident in the pairwise
relations \(\{e_{T} \perp\!\!\!\perp e_{X}\}\),
\(\{e_{T} \perp\!\!\!\perp e_{Y}\}\), and
\(\{e_{X} \perp\!\!\!\perp e_{Y}\}\), as well as colliders such as
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\). The
symbol \(\perp\!\!\!\perp\) denotes a concept know as
\emph{d-separation}, which roughly implies the independence of the
variables. This concept will be introduced in the next section.
